---
title: "Download transit data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download transit data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(GTFShift)

library(tidytransit)
library(sf)
library(mapview)
```

# Introduction

Performing analysis over transit requires data on the operation and infrastructure. GTFShift includes some built in methods that can you assist gathering it.

# Download GTFS files 

## Use open catalogues 

The gathering of the GTFS files can be simplified by using public archives like [mobilitydatabase.org](https://mobilitydatabase.org/) or [transit.land](https://www.transit.land/).

GTFShift provides a method to query Mobility Database: `GTFShift::query_mobilitydatabase()`. It queries the `/v1/gtfs_feeds` API endpoint, returning a list of GTFS feeds with information about the providers, the area they cover and an URL to download them.

To use it, an access token must be provided. It can be obtained for free at Mobility Database website.

## Use GTFShift incorporated database for Portugal

This library offers a small database with a compilation of GTFS files for Portuguese operators. It is a CSV file, available at `extdata/gtfs_sources_pt.csv`, and has the following attributes:

- `ID`, a string unique identifier of the region/city the GTFS file applies to;
- `LastUpdate`, the date at which this database entry was last updated;
- `ReferenceDate`, a representative Wednesday that falls within the GTFS calendar;
- `URL`, the URL at which the GTFS file is available;
- `GTFSDocs`, the URL to the page that documents the operator GTFS.


```{r message=TRUE, warning=FALSE}
data <- read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
summary(data)
data[1:4,]
```

## Download and validate integrity

`GTFShift` library offers a method to download and fix any integrity violations at GTFS files: `GTFShift::download_feed()`. It fixes inconsistencies on the `stop_times.txt` and missing `shapes.txt` files and returns a tidygtfs object, that can be used in any other method of GTFShift or tidyrtransit packages.

```{r message=TRUE, warning=TRUE}
# DOWNLOAD GTFS and store it locally
gtfs <- GTFShift::download_feed(data[data$ID=="cp",]$URL)
summary(gtfs)
```

# Download bus lanes 

Bus lanes can improve bus transit operation. Understanding their spatial distribution is important to study operation dynamics. `GTFShift::download_bus_lanes` allows to obtain the bus lanes network on Open Street Maps for a given area.

```{r}
aml <- st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg")
aml <- st_transform(aml, 4326)

lisboa <- aml[aml$Concelho == "Lisboa", ]

bbox = st_bbox(lisboa)

bus_lanes <- GTFShift::download_bus_lanes(bbox)

mapview::mapview(bus_lanes, layer.name = "Bus lanes")
```

