---
title: "1. Getting transit data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting transit data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  comment = "#>",
  engine.path = list(python = '~/.virtualenvs/r-reticulate/bin/python3.12')
)
```

```{r setup, message=FALSE, cache=FALSE}
library(GTFShift)
library(tidytransit)
```

# Introduction

Performing analysis over transit requires data on the operation and infrastructure. [GTFS](https://gtfs.org/) is an open standard that offers this information in a simple and complete way. 

GTFShift includes some built in methods that can assist gathering it, explored in this article.

# Read a GTFS file

GTFShift makes use of several R packages to implement its features, but most of it is built upon [tidytransit](https://r-transit.github.io/tidytransit/), so all the methods that accept a GTFS object should be a `tidygtfs` object.

To create it, you can either use `GTFShift::load_feed()` or the tidytransit method, `tidytransit::read_gtfs()`. Actually, `GTFShift::load_feed()` is an extension of the last, with the additional features of scanning the feed for any integrity errors and fixing them automatically, as well as the option to store it locally.

```{r eval = FALSE}
# Using tidytransit
gtfs = tidytransit::load_feed("https://operator.com/gtfs.zip")

# Using GTFShift
gtfs = GTFShift::load_feed("https://operator.com/gtfs.zip")
```


# Find GTFS feeds

## Using open catalogues 

The gathering of the GTFS files can be simplified by using public archives like [mobilitydatabase.org](https://mobilitydatabase.org/) or [transit.land](https://www.transit.land/).

GTFShift provides a method to query Mobility Database: `GTFShift::query_mobilitydatabase()`.
It queries the `/v1/gtfs_feeds` API endpoint, returning a list of GTFS feeds with information about the providers, the area they cover and an URL to download them.

To use it, an access token must be provided. It can be obtained for free at Mobility Database [website](https://mobilitydatabase.org/account).

```{r}
aml = sf::st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg", quiet = TRUE) |> sf::st_bbox()

# usethis::edit_r_environ() # to set MOBILITY_DATABASE variable for this code chunk to work
feeds = GTFShift::query_mobilitydatabase(
  refresh_token = Sys.getenv("MOBILITY_DATABASE"),
  bounding_filter_method = "completely_enclosed",
  bbox = aml
)

feeds |>
  dplyr::filter(status == "active") |>
  dplyr::select(provider, status, producer_url) |>
  head()

gtfs = GTFShift::load_feed(feeds$producer_url[2], create_transfers=FALSE)
summary(gtfs)
```


## Using GTFShift incorporated database for Portugal

This library offers a small database with a compilation of GTFS files for Portuguese operators. It is a CSV file, available at `extdata/gtfs_sources_pt.csv`, and has the following attributes:

- `ID`, a string unique identifier of the region/city the GTFS file applies to;
- `LastUpdate`, the date at which this database entry was last updated;
- `ReferenceDate`, a representative Wednesday that falls within the GTFS calendar;
- `URL`, the URL at which the GTFS file is available;
- `GTFSDocs`, the URL to the page that documents the operator GTFS.


```{r message=TRUE, warning=FALSE}
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
head(data)

gtfs = GTFShift::load_feed(data$URL[1], create_transfers = FALSE) # example with CP - trains
summary(gtfs)
```


