---
title: "Filter GTFS feeds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filter GTFS feeds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(GTFShift)

library(tidytransit)
library(dplyr)
library(mapview)
```


# Introduction 

GTFS feeds do not have a defined scope regarding its coverage of the transportation system. Some can be bounded to one agency, whereas others can aggregate several modes in the same city, or even national wise.

From the simpler to the most complex feeds, some analysis require to narrow the perspective. GTFShift provides some to help in this process.



# Filters

## Filter by agency

> This example uses a national GTFS feed for long distance rail in Germany, retrieved from https://gtfs.de/en/feeds/de_fv/.

```{r}
# Load GTFS
gtfs_local <- GTFShift::download("https://download.gtfs.de/germany/fv_free/latest.zip", "database/transit/germany_gtfs.zip")

gtfs <- tidytransit::read_gtfs(gtfs_local)
summary(gtfs)
```

Multimodal feeds aggregate several agencies. `GTFShift::filter_by_agency()`, allows to filter by the agency id, name, or both. You can filter by the `id` and/or `name` of the agency.

```{r}
# Filter by agency id
gtfs_5 <- GTFShift::filter_by_agency(gtfs, id=5)
summary(gtfs_5)

# Filter by agency name
gtfs_sncf <- GTFShift::filter_by_agency(gtfs, name="SNCF")
summary(gtfs_sncf)
```

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### Original GTFS
```{r}
shape_agency <- gtfs$trips %>%
  left_join(gtfs$routes, by = "route_id") %>%
  left_join(gtfs$agency, by = "agency_id") %>%
  select(shape_id, agency_id, agency_name) %>%
  distinct()
shape_agency$name = paste(shape_agency$agency_id, shape_agency$agency_name, sep=": ")
shapes_sf = tidytransit::shapes_as_sf(gtfs$shapes) %>%
  left_join(shape_agency, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "name", legend = TRUE)
```
</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### GTFS filtered for agency `id=5`
```{r}
shape_agency <- gtfs_5$trips %>%
  left_join(gtfs_5$routes, by = "route_id") %>%
  left_join(gtfs_5$agency, by = "agency_id") %>%
  select(shape_id, agency_id, agency_name) %>%
  distinct()
shape_agency$name = paste(shape_agency$agency_id, shape_agency$agency_name, sep=": ")
shapes_sf = tidytransit::shapes_as_sf(gtfs_5$shapes) %>%
  left_join(shape_agency, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "name", legend = TRUE)
```
</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### GTFS filtered for agency `name=SNCF`
```{r}
shape_agency <- gtfs_sncf$trips %>%
  left_join(gtfs_sncf$routes, by = "route_id") %>%
  left_join(gtfs_sncf$agency, by = "agency_id") %>%
  select(shape_id, agency_id, agency_name) %>%
  distinct()
shape_agency$name = paste(shape_agency$agency_id, shape_agency$agency_name, sep=": ")
shapes_sf = tidytransit::shapes_as_sf(gtfs_sncf$shapes) %>%
  left_join(shape_agency, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "name", legend = TRUE)
```
</div>
</div>


## Filter by mode


> This example uses a metropolitan wide GTFS feed for Los Angeles rail services, retrieved from https://developer.metro.net/gtfs-schedule-data/.

```{r}
# Load GTFS
gtfs_local <- GTFShift::download("https://gitlab.com/LACMTA/gtfs_rail/raw/master/gtfs_rail.zip", "database/transit/la_gtfs.zip")

gtfs <- tidytransit::read_gtfs(gtfs_local)
summary(gtfs)
```

Multimodal feeds aggregate several modes. `GTFShift::filter_by_mode()` allows to restrict the feed to some modes only. Refer to routes.txt 'route_type' parameter on [GTFS documentation](https://gtfs.org/documentation/schedule/reference/#routestxt) for more details on the modes id that should be used as parameters.

```{r}
# Filter by mode tram
gtfs_tram <- GTFShift::filter_by_modes(gtfs, modes=list(0))
summary(gtfs_tram)
```

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### Original GTFS
```{r}
shape_route <- gtfs$trips %>%
  left_join(gtfs$routes, by = "route_id") %>%
  select(shape_id, route_id, route_type) %>%
  distinct()
shape_route$route_type = as.character(shape_route$route_type)
shapes_sf = tidytransit::shapes_as_sf(gtfs$shapes) %>%
  left_join(shape_route, by = "shape_id") %>%
  filter(!is.na(route_type))
mapview::mapview(shapes_sf, zcol = "route_type", legend = TRUE)
```
</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### GTFS filtered for mode tram (0)
```{r}
shape_route <- gtfs_tram$trips %>%
  left_join(gtfs_tram$routes, by = "route_id") %>%
  select(shape_id, route_id, route_type) %>%
  distinct()
shape_route$route_type = as.character(shape_route$route_type)
shapes_sf = tidytransit::shapes_as_sf(gtfs_tram$shapes) %>%
  left_join(shape_route, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "route_type", legend = TRUE)
```
</div>
</div>


### Filter by route_name

> This article uses a GTFS feed from the library GTFS database for Portugal as an example. Refer to the \code{vignette("download")} for more details.

```{r}
# Get GTFS from library GTFS database for Portugal
data <- read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_local <- GTFShift::download(data[data$ID=="faro",]$URL, "database/transit/faro_gtfs.zip")

gtfs <- tidytransit::read_gtfs(gtfs_local)
summary(gtfs)
```

GTFS feeds aggregate several routes. `GTFShift::filter_by_route_name()`, allows to filter the feed for specific routes, given a partial or total match with the short or the long name.

```{r}
# Filter by short_name with exact match
gtfs_1 <- GTFShift::filter_by_route_name(gtfs, values=list("1"), short_name=TRUE, exact_match=TRUE)
summary(gtfs_1)

# Filter by long_name with partial match
gtfs_terminal <- GTFShift::filter_by_route_name(gtfs, values=list("Terminal", "Rodoviário"), short_name=FALSE, exact_match=FALSE)
summary(gtfs_terminal)
```

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### Original GTFS
```{r}
route_name <- gtfs$trips %>%
  left_join(gtfs$routes, by = "route_id") %>%
  select(shape_id, route_id, route_short_name, route_long_name) %>%
  distinct()
route_name$name = paste(route_name$route_short_name, route_name$route_long_name, sep=": ")
shapes_sf = tidytransit::shapes_as_sf(gtfs$shapes) %>%
  left_join(route_name, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "name", legend = TRUE)
```
</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### GTFS filtered for route short name `1` (exact match)
```{r}
route_name <- gtfs_1$trips %>%
  left_join(gtfs_1$routes, by = "route_id") %>%
  select(shape_id, route_id, route_short_name, route_long_name) %>%
  distinct()
route_name$name = paste(route_name$route_short_name, route_name$route_long_name, sep=": ")
shapes_sf = tidytransit::shapes_as_sf(gtfs_1$shapes) %>%
  left_join(route_name, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "name", legend = TRUE)
```
</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">
### GTFS filtered for route long name `Terminal`, `Rodoviário` (partial match)
```{r}
route_name <- gtfs_terminal$trips %>%
  left_join(gtfs_terminal$routes, by = "route_id") %>%
  select(shape_id, route_id, route_short_name, route_long_name) %>%
  distinct()
route_name$name = paste(gtfs_terminal$route_short_name, route_name$route_long_name, sep=": ")
shapes_sf = tidytransit::shapes_as_sf(gtfs_terminal$shapes) %>%
  left_join(route_name, by = "shape_id")
mapview::mapview(shapes_sf, zcol = "name", legend = TRUE)
```
</div>
</div>




