---
title: "5. Get OSM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Get OSM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  comment = "#>",
  engine.path = list(python = '~/.virtualenvs/r-reticulate/bin/python3.12')
)
```

```{r setup, message=FALSE, cache=FALSE}
library(GTFShift)
library(tidytransit)
```

# Introduction

OpenStreetMaps (OSM) is an important data source for transit analysis, due to its rich, open, and detailed geographic data.
GTFShift includes some methods that allow to access its information directly.


# Download bus lanes 

Bus lanes can improve bus transit operation. Understanding their spatial distribution is important to study operation dynamics.
`osm_bus_lanes` allows to obtain the bus lanes network on OpenStreetMaps for a given area.

```{r}
aml = sf::st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg", quiet = TRUE)
lisboa = aml |> dplyr::filter(Concelho == "Lisboa") |> sf::st_bbox()

bus_lanes = GTFShift::osm_bus_lanes(lisboa)
```
```{r, cache=FALSE}
mapview::mapview(bus_lanes, layer.name = "Bus lanes")
```



# Get centerlines for OSM road network

Performing an aggregated analysis for the spatial distribution of the route frequencies might be easier if the routes are projected over a simplified road network (refer to [vignette("analyse")](./analyse.html#improve-visualization-with-network_overline) for more details).

`osm_centerlines()` allows to generate this simplification by creating the centerlines for the road network exported from OpenStreetMaps, using Python [neatnet](https://uscuni.org/neatnet/) package.

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Original network

```{r original_network}
library(osmdata)

road_osm = opq("Arroios, Lisboa, Portugal") |>
    add_osm_feature(key = "highway", value = c("motorway", "trunk", "primary", "secondary", "tertiary", "residential", "unclassified", "living_street")) |>
    add_osm_feature(key = "area", value = "!yes") |>
    osmdata_sf() |>
    osm_poly2line()

road_osm = road_osm$osm_lines
```
```{r, cache=FALSE}
mapview::mapview(road_osm)
```


</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Simplified network

```{r simplified_network}
centerlines = GTFShift::osm_centerlines(place="Arroios, Lisboa, Portugal")
```
```{r, cache=FALSE}
mapview::mapview(centerlines)
```

</div>
</div>

