---
title: "6. Get OSM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Get OSM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  comment = "#>",
  engine.path = list(python = "~/.virtualenvs/r-reticulate/bin/python3.12")
)
```

```{r setup, message=FALSE, cache=FALSE}
library(GTFShift)
library(tidytransit)
library(dplyr)
```

# Introduction

OpenStreetMaps (OSM) is an important data source for transit analysis, due to its rich, open, and detailed geographic data.
GTFShift includes some methods that allow to access its information directly.

> Mind that despite this tutorial is mostly focused on the bus mode, GTFShift supports other transit modes, through the parameter `osm_route_type`, available in most methods that deal with OSM data. Refer to each method's documentation for more details.

## Using OSM API vs. local OSM extract

By default, GTFShift methods use OSM API to access OpenStreetMaps data ([osmdata::osmdata_sf](https://docs.ropensci.org/osmdata/reference/osmdata_sf.html)). 
However, the API rate is limited, making it hard to download large datasets. 
In these cases, it is recommended to use a local OSM extract file.
For this purpose, most of the methods support the optional parameter `osm_file`. Check each method documentation
to confirm.

```{r, eval=FALSE}
# Example of using local OSM extract

# Download OSM extract
# (use day tag on filename to get the latest file, as oe_download caches files and GeoFrabrick is updated daily)
osm_file <- oe_download(
  "https://download.geofabrik.de/europe/portugal-latest.osm.pbf",
  file_basename = sprintf("%s_%s.osm.pbf", "PT", format(Sys.Date(), "%Y%m%d"))
)

# Get GTFS from library GTFS database for Portugal
data <- read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id <- "lisboa"
gtfs <- GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers = FALSE)

# Build OSM query
library(osmdata)
q <- opq("Lisbon") |>
  add_osm_feature(key = "route", value = c("bus")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)

# Call method with osm_file parameter
shapes_ways_osm <- GTFShift::osm_shapes_to_routes(gtfs, q, ways = TRUE, osm_file = osm_file)
```



Refer to [osmextract::oe_download](https://docs.ropensci.org/osmextract/reference/oe_download.html) for a detailed
download documentation.

## OSM API instances

Another alternative to avoid OSM API usage rate limits is to use an alternative instance. There are several documented at [OSM Wiki](https://wiki.openstreetmap.org/wiki/Overpass_API#Public_Overpass_API_instances). Below is an example of how to change the instance to Private.coffee Overpass Instance.

```{r, eval=FALSE}
library(osmdata)

get_overpass_url() # You will see the default API instance
set_overpass_url("https://overpass.private.coffee/api/interpreter") # Changin the instance to Private.coffee
get_overpass_url() # You should see the new instance now

# Now you can call any method on osmdata and the new API instance will be used 
```

# Download bus lanes 

Dedicated bus lanes can improve bus transit operation. Understanding their spatial distribution is important to study operation dynamics.
`osm_bus_lanes` allows to obtain the bus lanes network on OpenStreetMaps for a given area.

```{r, eval=FALSE}
aml <- sf::st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg", quiet = TRUE)
lisboa <- aml |>
  dplyr::filter(Concelho == "Lisboa") |>
  sf::st_bbox()

bus_lanes <- GTFShift::osm_bus_lanes(lisboa) |> select(osm_id)
```

```{r include=FALSE}
bus_lanes <- sf::st_read("https://github.com/U-Shift/busclar/releases/download/0.1/bus_lanes_osm_lisbon.gpkg", quiet = TRUE)
```

```{r, cache=FALSE}
mapview::mapview(bus_lanes, layer.name = "Bus lanes")
```


# Get OSM data for transit routes

OpenStreetMaps defines transit routes as a [relation](https://wiki.openstreetmap.org/wiki/Public_transport#Service_routes) of ways (roads or railways, for instance) and nodes (stops and platforms). GTFShift provides methods to use them in the GTFS analysis.

> For demonstration purposes, the next snippets will use the Lisbon urban bus network.

```{r eval=FALSE}
# Get GTFS from library GTFS database for Portugal
data <- read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id <- "lisboa"
gtfs <- GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers = FALSE)

# Build OSM query
library(osmdata)
q <- opq("Lisbon") |>
  add_osm_feature(key = "route", value = c("bus")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)
```


## Routes (matching GTFS id)

GTFS routes shapes and OSM transit routes are linked through OSM `gtfs:*` keys. `GTFShift::osm_shapes_to_routes()` allows to query OSM for the routes matching the feed trips, given their shape id.

```{r include=FALSE}
shapes_geometry_osm <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.7.0/osm_carris_shapes_to_routes_794.gpkg")
```


```{r eval=FALSE}
# Subset feed for some routes only, for demonstration purposes
gtfs_794 <- GTFShift::filter_by_route_name(gtfs, list("794"))

# Match shapes geometry
shapes_geometry_osm <- GTFShift::osm_shapes_to_routes(gtfs_794, q)
```

```{r}
shapes_geometry_osm
```

```{r include=FALSE}
shapes_sf <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.7.0/osm_carris_shapes_to_routes_794_original.gpkg")
```

```{r eval=FALSE}
# Get original shapes, for comparison
shapes_sf <- tidytransit::shapes_as_sf(gtfs_794$shapes)
```




<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### GTFS shapes

```{r, cache=FALSE}
mapview::mapview(shapes_sf, zcol = "shape_id", legend = TRUE, layer.name = "GTFS shapes")
```

</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### OSM routes

```{r, cache=FALSE}
mapview::mapview(shapes_geometry_osm, zcol = "shape_id", legend = TRUE, layer.name = "OSM routes")
```

</div>
</div>

## Ways (matching GTFS id)

By setting the `ways` parameter to `TRUE`, `GTFShift::osm_shapes_to_routes()` enables the retrieval of the OSM routes disaggregated by the individual ways that compose them.

```{r include=FALSE}
shapes_ways_osm <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.7.0/osm_carris_shapes_to_routes_794_ways.gpkg")
```


```{r eval=FALSE}
# Match shapes geometry disaggregated by ways
shapes_ways_osm <- GTFShift::osm_shapes_to_routes(gtfs_794, q, ways = TRUE)
```

```{r}
shapes_ways_osm |> select(shape_id, osm_id, way_osm_id, lanes)
```


```{r, cache=FALSE}
mapview::mapview(shapes_ways_osm, zcol = "way_osm_id", legend = FALSE, layer.name = "OSM ways")
```

## Routes (matching shapes geometry)

Despite existing, OSM `gtfs:*` keys are not widely used. In July 2025, only 3.1% of relations tagged as `route=bus` had the `gtfs:shape_id` key set (9 784 of 312 049).

To overcome this issue, `GTFShift::osm_shapes_match_routes()` performs the association between the OSM transit routes and the GTFS shapes considering a geometrical match. For each GTFS route identifier (provided as a parameter), the function first selects the subset of OSM route relations that match it. Then, for each GTFS shape associated with the route, it identifies the most similar OSM route relation by minimizing a cost function that combines the distance between start and end points, total length, and number of stops.

> The match between the GTFS and OSM identifiers is strict by default, but parameter `gtfs_osm_match_exact` can be set to `FALSE` to allow a partial match (e.g., matching GTFS route `15E` with OSM routes `Carris/15E`, `15E-1`, etc.).

> GTFShift provides a test script that runs this method for several cities around the world at [dev/test_osm_shapes_match_routes_local.R](https://github.com/U-Shift/GTFShift/blob/main/dev/test_osm_shapes_match_routes_local.R).

```{r include=FALSE}
shapes_match_routes <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.7.0/osm_carris_shapes_match_routes_736_750_15E_65B.gpkg")
```


```{r eval=FALSE}
# Subset feed for some routes only, for demonstration purposes
gtfs_subset <- GTFShift::filter_by_route_name(gtfs, list("736", "750", "15E", "65B"))

# Match shapes geometry
shapes_match_routes <- GTFShift::osm_shapes_match_routes(gtfs_subset, q)
#> > Found 14 GTFS shapes and 231 stops
#> > Found 300 OSM route relations and 4911 bus stops/platforms
#> > Associated 14 shapes (100.00% of 14 total) of 8 routes (100.00% of 8 total) with OSM routes, with a mean distance of 25.42 meters for points, 49.86 meters for route length and a mean difference of 0.43 stops
#> > Of those, 14 shapes (100.00% of 14 matched) have a distance difference below 1000 meters, a points difference below 500 meters
```

```{r}
summary(shapes_match_routes)

# Visualize results
shapes_match_routes$map_name <- paste(
  shapes_match_routes$route_short_name,
  " | ",
  shapes_match_routes$shape_id,
  " | ",
  shapes_match_routes$osm_id
)
```


```{r, cache=FALSE}
mapview::mapview(shapes_match_routes, zcol = "map_name", legend = TRUE, layer.name = "route_short_name | shape_id | osm_id")
```


### Updating OSM routes with shape_id

The association between OSM route relations id and the GTFS shapes_id returned by `GTFShift::osm_shapes_match_routes()` can be used to update OpenStreetMaps data. Refer to [Update OSM data](./osm_update.html) for more details.

### Validating OSM routes

The warning logs displayed by `GTFShift::osm_shapes_match_routes()` can be used to validate the OSM routes, identifying shapes without a match, or even OSM existing routes that violate OSM integrity rules, such as having stops in the wrong order.


# Get centerlines for OSM road network

Performing an aggregated analysis for the spatial distribution of the route frequencies might be easier if the routes are projected over a simplified road network (refer to [vignette("analyse")](./analyse.html#improve-visualization-with-network_overline) for more details).

`GTFShift::osm_centerlines()` allows to generate this simplification by creating the centerlines for the road network exported from OpenStreetMaps, using Python [neatnet](https://uscuni.org/neatnet/) package.

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Original network

```{r include=FALSE}
road_osm <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.7.0/osm_arroios.gpkg")
```


```{r eval=FALSE}
library(osmdata)

road_osm <- opq("Arroios, Lisboa, Portugal") |>
  add_osm_feature(key = "highway", value = c("motorway", "trunk", "primary", "secondary", "tertiary", "residential", "unclassified", "living_street")) |>
  add_osm_feature(key = "area", value = "!yes") |>
  osmdata_sf() |>
  osm_poly2line()

road_osm <- road_osm$osm_lines
```
```{r, cache=FALSE}
mapview::mapview(road_osm)
```


</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Simplified network

```{r include=FALSE}
centerlines <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.7.0/osm_arroios_centerlines.gpkg")
```


```{r eval=FALSE}
centerlines <- GTFShift::osm_centerlines(place = "Arroios, Lisboa, Portugal")
#> Using Python: /usr/bin/python3.12
#> Creating virtual environment '~/.virtualenvs/r-reticulate' ...
#> Done!
#> Installing packages: pip, wheel, setuptools
#> Installing packages: numpy
#> Virtual environment '~/.virtualenvs/r-reticulate' successfully created.
#> Using virtual environment '/home/runner/.virtualenvs/r-reticulate' ...
```
```{r, cache=FALSE}
mapview::mapview(centerlines)
```

</div>
</div>

