---
title: "5. Get OSM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Get OSM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  comment = "#>",
  engine.path = list(python = '~/.virtualenvs/r-reticulate/bin/python3.12')
)
```

```{r setup, message=FALSE, cache=FALSE}
library(GTFShift)
library(tidytransit)
```

# Introduction

OpenStreetMaps (OSM) is an important data source for transit analysis, due to its rich, open, and detailed geographic data.
GTFShift includes some methods that allow to access its information directly.


# Download bus lanes 

Bus lanes can improve bus transit operation. Understanding their spatial distribution is important to study operation dynamics.
`osm_bus_lanes` allows to obtain the bus lanes network on OpenStreetMaps for a given area.

```{r}
aml = sf::st_read("https://github.com/U-Shift/MQAT/raw/refs/heads/main/geo/MUNICIPIOSgeo.gpkg", quiet = TRUE)
lisboa = aml |> dplyr::filter(Concelho == "Lisboa") |> sf::st_bbox()

bus_lanes = GTFShift::osm_bus_lanes(lisboa)
```
```{r, cache=FALSE}
mapview::mapview(bus_lanes, layer.name = "Bus lanes")
```



# Get centerlines for OSM road network

Performing an aggregated analysis for the spatial distribution of the route frequencies might be easier if the routes are projected over a simplified road network (refer to [vignette("analyse")](./analyse.html#improve-visualization-with-network_overline) for more details).

`osm_centerlines()` allows to generate this simplification by creating the centerlines for the road network exported from OpenStreetMaps, using Python [neatnet](https://uscuni.org/neatnet/) package.

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Original network

```{r original_network}
library(osmdata)

road_osm = opq("Arroios, Lisboa, Portugal") |>
    add_osm_feature(key = "highway", value = c("motorway", "trunk", "primary", "secondary", "tertiary", "residential", "unclassified", "living_street")) |>
    add_osm_feature(key = "area", value = "!yes") |>
    osmdata_sf() |>
    osm_poly2line()

road_osm = road_osm$osm_lines
```
```{r, cache=FALSE}
mapview::mapview(road_osm)
```


</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Simplified network

```{r simplified_network}
centerlines = GTFShift::osm_centerlines(place="Arroios, Lisboa, Portugal")
```
```{r, cache=FALSE}
mapview::mapview(centerlines)
```

</div>
</div>


# Get OSM bus routes

OpenStreetMaps defines bus routes as a [relation](https://wiki.openstreetmap.org/wiki/Tag:route%3Dbus) of ways (usually roads) and nodes (stops and platforms). GTFShift provides methods to use them in the GTFS analysis.

> For demonstration purposes, the next snippets will use the Lisbon urban bus network.

```{r}
# Get GTFS from library GTFS database for Portugal
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id = "lisboa"
gtfs = GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers=FALSE)

# Build OSM query
library(osmdata)
q = opq("Lisbon")  |>
  add_osm_feature(key = "route", value = c("bus", "tram")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)
```


## Matching GTFS ids

GTFS routes shapes and OSM bus routes are linked through OSM `gtfs:*` keys. `GTFShift::osm_shapes_to_routes()` and `GTFShift::osm_trips_to_routes()` allow to query OSM for the routes matching the feed trips, given, respectively, their shape or trip id.

```{r}
# Subset feed for some routes only, for demonstration purposes
gtfs_794 = GTFShift::filter_by_route_name(gtfs, list("794"))

# Match shapes geometry
shapes_geometry_osm = GTFShift::osm_shapes_to_routes(gtfs_794, q)

shapes_geometry_osm
```


<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### GTFS shapes

```{r, cache=FALSE}
shapes_sf = tidytransit::shapes_as_sf(gtfs_794$shapes)
mapview::mapview(shapes_sf, zcol = "shape_id", legend = TRUE, layer.name="GTFS shapes")
```

</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### OSM routes

```{r, cache=FALSE}
mapview::mapview(shapes_geometry_osm, zcol = "shape_id", legend = TRUE, layer.name="OSM routes")
```

</div>
</div>

## Matching shapes geometry

Despite existing, OSM `gtfs:*` keys are not widely used. In July 2025, only 3.1% of relations tagged as `route=bus` had the `gtfs:shape_id` key set (9 784 of 312 049).

To overcome this issue, `GTFShift::osm_shapes_match_routes()` performs the association between the OSM bus routes and the GTFS shapes considering, for each route, the start and end points and total length.

```{r}
# Subset feed for some routes only, for demonstration purposes
gtfs_subset = GTFShift::filter_by_route_name(gtfs, list("736", "750", "15E", "65B"))

# Match shapes geometry
shapes_match_routes = GTFShift::osm_shapes_match_routes(gtfs_subset, q)

summary(shapes_match_routes)
```


```{r, cache=FALSE}
# Visualize results
shapes_match_routes$map_name = paste(
  shapes_match_routes$route_short_name,
  " // ",
  shapes_match_routes$shape_id, 
  " // ", 
  shapes_match_routes$osm_id
)
mapview::mapview(shapes_match_routes, zcol = "map_name")
```

### Updating OSM routes with shape_id

The association between OSM route relations id and the GTFS shapes_id returned by `GTFShift::osm_shapes_match_routes()` can be used to update OpenStreetMaps dataset.

Python library [OsmApi](https://osmapi.metaodi.ch/osmapi/OsmApi.html) enables to perform this batch process using OpenStreetMaps API. GTFShift provides a script to perform this operation using OsmApi. Check it at the repository, at `inst/python/OSM_routes_add_shapes.py`, or find its location on your computer, using `system.file("python", "osm_routes_add_shapes.py", package = "GTFShift")`.
