---
title: "Extra. Update OSM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extra. Update OSM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  eval = FALSE, # Don't run code, just display!
  comment = "#>",
  engine.path = list(python = "~/.virtualenvs/r-reticulate/bin/python3.12")
)
```

# Introduction

The association between OSM route relations id and the GTFS shapes_id returned by `GTFShift::osm_shapes_match_routes()` can be used to update OpenStreetMaps data.

Python library [OsmApi](https://osmapi.metaodi.ch/osmapi/OsmApi.html) enables to perform this batch process using OpenStreetMaps API. This article, adapted from an [osmapi example script](https://github.com/metaodi/osmapi/blob/16aeb189c9ff8db607cc118842133b8cd7b60971/examples/oauth2.py), aims to document this procedure.

It can be run either at the development or live version of OSM, requiring credentials to be run on either of them.

This script considers credentials to be stored at environment variables (`OSM_USER` and `OSM_PASSWORD`), to avoid hardcoding them in the script. When running it, either replace them manually by your credentials or, better, set them as environment variables, using `usethis::edit_r_environ()` on R, or by exporting them in the terminal.

# Match GTFS shapes with OSM routes 

This example assumes that the steps described at [Matching shapes geometry](./osm.html#matching-shapes-geometry) have been followed, up to the definition of the variable `shapes_match_routes`. To use it in Python, it should be stored in a CSV file.

```{r}
write.csv(shapes_match_routes |> sf::st_drop_geometry() |> mutate(
  distance_diff = round(distance_diff),
  points_diff = round(points_diff)
), "osm_match.csv", row.names = FALSE)
```

Then, it should be loaded into the Python environment, filtering the matches by the criteria that best suits your context.
```{python}
import pandas as pd

area = "stcp"
gtfs_date = "2026-04-16"
run_date = "20260416"
base_folder = "../../GTFShift-web/scripts/osm_match" # "./" # "../../GTFShift-web/scripts/osm_match"

df = pd.read_csv(f"{base_folder}/{area}/shapes_match_stcp_gtfs{gtfs_date}_run{run_date}.csv") # CSV with columns osm_id, shape_id and route_id (optional)
len(df)
df = df[(df['distance_diff'] < 1000) & (df['points_diff'] < 500)] # Filter to only update those that meet threshold
len(df)
```


# Local environment set up

Load dependencies
```{python}
from requests_oauth2client import OAuth2Client, OAuth2AuthorizationCodeAuth
import requests
import webbrowser
import osmapi
from dotenv import load_dotenv, find_dotenv
import os
import pandas as pd
import logging
```

If there is any dependency missing, just install it with `reticulate::py_install()`.
```{r}
# library(reticulate)
# py_install("dotenv")
```

Load environment variables
```{python}
load_dotenv(find_dotenv())
```

Set up logging
```{python}
# If previous logger definition exists, remove it to avoid duplicated logs
if 'logger' in globals():
    logger.handlers.clear()
logger = logging.getLogger('my_logger')
logger.setLevel(logging.DEBUG)  # Set the base logging level
log_file_path = f'{base_folder}/{area}/shapes_match_{area}_gtfs{gtfs_date}_run{run_date}.osm.log'
file_handler = logging.FileHandler(log_file_path)
file_handler.setLevel(logging.DEBUG)
console_handler = logging.StreamHandler()
console_handler.setLevel(logging.INFO)
formatter = logging.Formatter('%(asctime)s - %(levelname)s - %(message)s')
file_handler.setFormatter(formatter)
console_handler.setFormatter(formatter)
logger.addHandler(file_handler)
logger.addHandler(console_handler)
```



# Connect to OSM

Open the API connection
```{python}
api_url = "https://api.openstreetmap.org" # Live OSM
# api_url = "https://api06.dev.openstreetmap.org" # Use this for testing on the development OSM instance, but make sure to change credentials accordingly

api = osmapi.OsmApi(api=api_url, username=os.getenv("OSM_USER"), password=os.getenv("OSM_PASSWORD"))
```


# Update OSM data

Attention! The script below will open a changeset and update OSM data! Make sure you have the right data before proceeding!

```{python}
# Create change set, updating relations with tag gtfs:shape_id
# The changeset comment can be customized to better describe the change submitted 
with api.Changeset({"comment": "GTFS shapes and routes association (using GTFShift v0.7.0)", "review_requested": "no", "locale": "pt", "source": "local knowledge"}) as changeset_id:
  logger.info(f"Running changeset {changeset_id} for {len(df)} relations")
  logger.info(f"{'route_id':20s}{'shape_id':20s} | {'osm_id':20s} | {'osm_route_id':20s}{'updated?':10s}{'osm_shape_id':20s}{'updated?':10s} | {'operation status':20s}")
  for idx, row in df.iterrows():
    route_id = str(row["route_id"]) if "route_id" in row else None
    shape_id = str(row["shape_id"])
    osm_id = int(row["osm_id"])
    
    relation = api.RelationGet(osm_id)
    relation_shape = str(relation["tag"]["gtfs:shape_id"]) if "gtfs:shape_id" in relation["tag"] else "-"
    relation_route = str(relation["tag"]["gtfs:route_id"]) if "gtfs:route_id" in relation["tag"] else "-"
    
    status = "-"
    if relation_shape != shape_id or route_id and (relation_route != route_id):
      relation["tag"]["gtfs:shape_id"] = str(shape_id) # https://wiki.openstreetmap.org/wiki/Key:gtfs:shape_id
      if route_id:
        relation["tag"]["gtfs:route_id"] = str(route_id) # https://wiki.openstreetmap.org/wiki/Key:gtfs:route_id
      update = api.RelationUpdate(relation)
      status = "Updated"
    else:
      status = "Skipped"
    
    logger.info(f"{str(route_id):20s}{str(shape_id):20s} | {osm_id:20d} | {str(relation_route):20s}{('⬜' if relation_route==route_id else '✏️'):10s}{str(relation_shape):20s}{('⬜' if relation_shape==shape_id else '✏️'):10s} | {status:20s}")
```


## Rollback (avoid this, please :/)

If you need to rollback the changes, use the code below.

```{python}
with api.Changeset({"comment": "GTFS shapes association rollback", "review_requested": "no", "locale": "pt", "source": "local knowledge"}) as changeset_id:
  for idx, row in df.iterrows():
    osm_id = int(row["osm_id"])
    relation = api.RelationGet(osm_id)
    relation_prev = api.RelationGet(osm_id, RelationVersion=relation["version"]-1)
    logger.info(f"{osm_id} {relation['tag']['gtfs:shape_id'] if 'gtfs:shape_id' in relation['tag'] else '-'} Current {relation['version']} Previous {relation_prev['version']}")
    relation_prev["version"] = relation["version"] # We need to set version to last to enable update
    update = api.RelationUpdate(relation_prev)
```

## Validate changes

```{python}
logger.info("Validating changes...")
logger.info(f"{'route_id':20s}{'shape_id':20s} | {'osm_id':20s} | {'osm_route_id':20s}{'osm_route_id==route_id':10s}{'osm_shape_id':20s}{'osm_shape_id==shape_id':10s}")
for idx, row in df.iterrows():
    shape_id = str(row["shape_id"])
    route_id = str(row["route_id"]) if "route_id" in row else None
    osm_id = int(row["osm_id"])
  
    relation = api.RelationGet(osm_id)
    
    relation_shape = str(relation["tag"]["gtfs:shape_id"]) if "gtfs:shape_id" in relation["tag"] else "-"
    relation_route = str(relation["tag"]["gtfs:route_id"]) if "gtfs:route_id" in relation["tag"] else "-"
  
    # if relation_shape!=shape_id or relation_route!=route_id: 
    logger.info(f"{route_id:20s}{shape_id:20s} | {osm_id:20d} | {relation_route:20s}{('✅' if relation_route==route_id else '❌'):10s}{relation_shape:20s}{('✅' if relation_shape==shape_id else '❌'):10s}")
logger.info("Validation finished!")
```



