---
title: "Extra. Update OSM data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extra. Update OSM data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  eval = FALSE, # Don't run code, just display!
  comment = "#>",
  engine.path = list(python = "~/.virtualenvs/r-reticulate/bin/python3.12")
)
```

# Introduction

The association between OSM route relations id and the GTFS shapes_id returned by `GTFShift::osm_shapes_match_routes()` can be used to update OpenStreetMaps data.

Python library [OsmApi](https://osmapi.metaodi.ch/osmapi/OsmApi.html) enables to perform this batch process using OpenStreetMaps API. This article, adapted from an [osmapi example script](https://github.com/metaodi/osmapi/blob/16aeb189c9ff8db607cc118842133b8cd7b60971/examples/oauth2.py), aims to document this procedure.

It can be run either at the development or live version of OSM, requiring API credentials to be run on either of them, that should be stored as environment variables.

> To get credentials for the development version, go to https://master.apis.dev.openstreetmap.org/oauth2/applications.
> They should be stored as `OSM_OAUTH_CLIENT_ID_DEV` and `OSM_OAUTH_CLIENT_SECRET_DEV`.

> To get credentials for the live version, go to https://www.openstreetmap.org/oauth2/applications.
> They should be stored as `OSM_OAUTH_CLIENT_ID_PROD` and `OSM_OAUTH_CLIENT_SECRET_PROD`.

> API registration parameters:
> Redirect uri: `urn:ietf:wg:oauth:2.0:oob`
> Permissions: `write_api`, `write_notes`, `read_prefs`

The environment variables can be edited using `usethis::edit_r_environ()`, on R.


# Match GTFS shapes with OSM routes 

This example assumes that the steps described at [Matching shapes geometry](./osm.html#matching-shapes-geometry) have been followed, up to the definition of the variable `shapes_match_routes`. To use it in Python, it should be stored in a CSV file.

```{r}
write.csv(shapes_match_routes |> sf::st_drop_geometry() |> mutate(
  distance_diff = round(distance_diff),
  points_diff = round(points_diff)
), "osm_match.csv", row.names = FALSE)
```

Then, it should be loaded into the Python environment, filtering the matches by the criteria that best suits your context.
```{python}
import pandas as pd

area = "stcp"
gtfs_date = "2026-04-16"
run_date = "20260416"
base_folder = "./"

df = pd.read_csv(f"{base_folder}/{area}/shapes_match_stcp_gtfs{gtfs_date}_run{run_date}.csv") # CSV with columns osm_id, shape_id and route_id (optional)
len(df)
df = df[(df['distance_diff'] < 1000) & (df['points_diff'] < 500)] # Filter to only update those that meet threshold
len(df)
```


# Local environment set up

Load dependencies
```{python}
from requests_oauth2client import OAuth2Client, OAuth2AuthorizationCodeAuth
import requests
import webbrowser
import osmapi
from dotenv import load_dotenv, find_dotenv
import os
import pandas as pd
import logging

# Patch requests default User-Agent to avoid being blocked by OSM (see https://github.com/Zverik/cli-oauth2/issues/7)
requests.utils.default_user_agent = lambda: "GTFShift/0.8.3 (ushift@tecnico.ulisboa.pt)"
```

If there is any dependency missing, just install it with `reticulate::py_install()`.
```{r}
# library(reticulate)
# py_install("geopy")
```

Load environment variables
```{python}
load_dotenv(find_dotenv())
```

Set up logging
```{python}
# If previous logger definition exists, remove it to avoid duplicated logs
if 'logger' in globals():
    logger.handlers.clear()
logger = logging.getLogger('my_logger')
logger.setLevel(logging.DEBUG)  # Set the base logging level
log_file_path = f'{base_folder}/{area}/shapes_match_{area}_gtfs{gtfs_date}_run{run_date}.osm.log'
file_handler = logging.FileHandler(log_file_path)
file_handler.setLevel(logging.DEBUG)
console_handler = logging.StreamHandler()
console_handler.setLevel(logging.INFO)
formatter = logging.Formatter('%(asctime)s - %(levelname)s - %(message)s')
file_handler.setFormatter(formatter)
console_handler.setFormatter(formatter)
logger.addHandler(file_handler)
logger.addHandler(console_handler)
```



# Connect to OSM

Either connect to the live OSM version...
```{python}
client_id = os.getenv("OSM_OAUTH_CLIENT_ID_PROD") # To edit env vars on RStudio, use usethis::edit_r_environ()
client_secret = os.getenv("OSM_OAUTH_CLIENT_SECRET_PROD")
authorization_base_url = "https://www.openstreetmap.org/oauth2/authorize"
token_url = "https://www.openstreetmap.org/oauth2/token"
api_url = "https://api.openstreetmap.org"
```

... or the dev one. It is recommended to try to apply the changes first at the dev version, to avoid corrupting the live OSM data.
```{python}
client_id = os.getenv("OSM_OAUTH_CLIENT_ID_DEV") # To edit env vars on RStudio, use usethis::edit_r_environ()
client_secret = os.getenv("OSM_OAUTH_CLIENT_SECRET_DEV")
authorization_base_url = "https://master.apis.dev.openstreetmap.org/oauth2/authorize"
token_url = "https://master.apis.dev.openstreetmap.org/oauth2/token"
api_url = "https://api06.dev.openstreetmap.org"
```

If no window is opened, authenticate by clicking on the url that is printed in the console.
```{python}
redirect_uri = "urn:ietf:wg:oauth:2.0:oob"
oauth2client = OAuth2Client(
    token_endpoint=token_url,
    authorization_endpoint=authorization_base_url,
    redirect_uri=redirect_uri,
    client_id=client_id,
    client_secret=client_secret,
    auth_method="client_secret_post",
    code_challenge_method=None
)

# Open OSM website to authorize user using the write_api and write_notes scope
scope = ["write_api", "write_notes", "read_prefs"]
az_request = oauth2client.authorization_request(scope=scope)
print(f"Authorize user using this URL: {az_request.uri}")
webbrowser.open(az_request.uri) # If on studio web, this might not work. If so, just open the link printed in the command before and jump to next line.
```

After authenticating, you will be presented with an authorization code. Copy it and set `auth_code` to its value.
```{python}
auth_code = "" # Replace with your authorization code
auth_code
auth = OAuth2AuthorizationCodeAuth(
    oauth2client,
    auth_code,
    redirect_uri=redirect_uri,
)
oauth_session = requests.Session()
oauth_session.auth = auth
```

Test authentication (should return 200 status code)
```{python}
resp = oauth_session.get(f"{api_url}/api/0.6/user/details")
print(resp.status_code) # 200 is expected
# print(resp.text) # This can help debugging if something goes wrong
```

Finally, open the API connection
```{python}
api = osmapi.OsmApi(api=api_url, session=oauth_session)
```


# Update OSM data

Attention! The script below will open a changeset and update OSM data! Make sure you have the right data before proceeding!

```{python}
# Create change set, updating relations with tag gtfs:shape_id
# The changeset comment can be customized to better describe the change submitted 
with api.Changeset({"comment": "GTFS shapes and routes association (using GTFShift v0.8.3)", "review_requested": "no", "locale": "pt", "source": "local knowledge"}) as changeset_id:
  logger.info(f"Running changeset {changeset_id} for {len(df)} relations")
  logger.info(f"{'route_id':20s}{'shape_id':20s} | {'osm_id':20s} | {'osm_route_id':20s}{'updated?':10s}{'osm_shape_id':20s}{'updated?':10s} | {'operation status':20s}")
  for idx, row in df.iterrows():
    route_id = str(row["route_id"]) if "route_id" in row else None
    shape_id = str(row["shape_id"])
    osm_id = int(row["osm_id"])
    
    relation = api.RelationGet(osm_id)
    relation_shape = str(relation["tag"]["gtfs:shape_id"]) if "gtfs:shape_id" in relation["tag"] else "-"
    relation_route = str(relation["tag"]["gtfs:route_id"]) if "gtfs:route_id" in relation["tag"] else "-"
    
    status = "-"
    if relation_shape != shape_id or route_id and (relation_route != route_id):
      relation["tag"]["gtfs:shape_id"] = str(shape_id) # https://wiki.openstreetmap.org/wiki/Key:gtfs:shape_id
      if route_id:
        relation["tag"]["gtfs:route_id"] = str(route_id) # https://wiki.openstreetmap.org/wiki/Key:gtfs:route_id
      update = api.RelationUpdate(relation)
      status = "Updated"
    else:
      status = "Skipped"
    
    logger.info(f"{str(route_id):20s}{str(shape_id):20s} | {osm_id:20d} | {str(relation_route):20s}{('⬜' if relation_route==route_id else '✏️'):10s}{str(relation_shape):20s}{('⬜' if relation_shape==shape_id else '✏️'):10s} | {status:20s}")
```


## Rollback (avoid this, please :/)

If you need to rollback the changes, use the code below.

```{python}
with api.Changeset({"comment": "GTFS shapes association rollback", "review_requested": "no", "locale": "pt", "source": "local knowledge"}) as changeset_id:
  for idx, row in df.iterrows():
    osm_id = int(row["osm_id"])
    relation = api.RelationGet(osm_id)
    relation_prev = api.RelationGet(osm_id, RelationVersion=relation["version"]-1)
    logger.info(f"{osm_id} {relation['tag']['gtfs:shape_id'] if 'gtfs:shape_id' in relation['tag'] else '-'} Current {relation['version']} Previous {relation_prev['version']}")
    relation_prev["version"] = relation["version"] # We need to set version to last to enable update
    update = api.RelationUpdate(relation_prev)
```

## Validate changes

```{python}
logger.info("Validating changes...")
logger.info(f"{'route_id':20s}{'shape_id':20s} | {'osm_id':20s} | {'osm_route_id':20s}{'osm_route_id==route_id':10s}{'osm_shape_id':20s}{'osm_shape_id==shape_id':10s}")
for idx, row in df.iterrows():
    shape_id = str(row["shape_id"])
    route_id = str(row["route_id"]) if "route_id" in row else None
    osm_id = int(row["osm_id"])
  
    relation = api.RelationGet(osm_id)
    
    relation_shape = str(relation["tag"]["gtfs:shape_id"]) if "gtfs:shape_id" in relation["tag"] else "-"
    relation_route = str(relation["tag"]["gtfs:route_id"]) if "gtfs:route_id" in relation["tag"] else "-"
  
    # if relation_shape!=shape_id or relation_route!=route_id: 
    logger.info(f"{route_id:20s}{shape_id:20s} | {osm_id:20d} | {relation_route:20s}{('✅' if relation_route==route_id else '❌'):10s}{relation_shape:20s}{('✅' if relation_shape==shape_id else '❌'):10s}")
logger.info("Validation finished!")
```



