---
title: "Prioritize bus lane implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prioritize bus lane implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  cache = TRUE,
  comment = "#>"
)
```

```{r setup, cache=FALSE}
library(GTFShift)
library(tidytransit)
library(mapview)
library(dplyr)
library(osmdata)
```

```{r include=FALSE}
Sys.sleep(60)
```

# Introduction

Bus lanes have the potential to improve the reliability of bus operations by limiting the negative impacts of traffic congestion, reducing the variability of travel times, and increasing the average commercial speed - which can ultimately be used to increase service frequency.

However, introducing them on road infrastructure affects other modes, such as private vehicles, which may experience a decrease in the level of service due to reduced allocated space, potentially jeopardizing public acceptance.

Common criteria for implementing bus lanes include:

-   **Frequency of buses (and trams) per hour and direction, at a peak hour;**
-   **Number of lanes in the same direction;**
-   **Existing traffic conditions**;
-   Existing bus lanes in the area (from a network continuity perspective).

GTFShift provides methods to analyse the dimensions in bold, namely:

-  `GTFShift::get_way_frequency_hourly()`, to obtain the frequency of services per hour for each road segment with transit service and its associated characteristics (such as number of lanes).

> `GTFShift::get_route_frequency_hourly()` is an alternative

- `GTFShift::osm_bus_lanes()`, to identify existing bus lanes in the road network.

- `GTFShift::rt_collect()`, to collect GTFS-RT data, which can be later used with `GTFShift::rt_extend_prioritization()` to include real-time operational metrics in the prioritization analysis.

This document explores how to use these methods in a combined way to assist public transport planners in prioritizing bus lane implementations. For details on the several encapsulated features, refer to the numerated articles in the menu, that explore in detail each of the specific approaches followed. 

# Prioritize lanes

## Generate base indicators

`GTFShift::prioritize_lanes()` is a simple method that generates indicators for most of the criteria mentioned above using GTFS and OpenStreetMaps data (service frequency and lane characteristics). With a single call, it returns a data.frame with the relevant metrics for each road segment with transit service. 

```{r}
# Get GTFS from library GTFS database for Portugal
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id = "lisboa"
gtfs = GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers=FALSE)

osm_q = opq(bbox=sf::st_bbox(tidytransit::shapes_as_sf(gtfs$shapes)))  |>
  add_osm_feature(key = "route", value = c("bus", "tram")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)

lanes = prioritize_lanes(gtfs, osm_q)
summary(lanes)
```


## Extend with GTFS-RT data

If GTFS-RT data is available, it can be used to extend the prioritization analysis with real-time operational metrics, such as average speed. This can help identify road segments where buses are experiencing significant delays due to traffic congestion, which may benefit from bus lane implementation.

Refer to the [GTFS Real Time](./rt.html) article for details on how to collect GTFS-RT data and extend the prioritization analysis.

## Visualize results

The aggregated data can then be manipulated according to the prioritization criteria defined by the user. For instance, the following code classifies (in red) the road segments considering a minimum frequency of 5 buses/hour and more than 1 lane per direction as high priority for bus lane implementation. 

> The classification of frequent service as 5 buses/hour is based on the HCM guidelines for level of service, where a frequency of 5 buses/hour corresponds to a level of service C or better. Refer to the [Classify transit data](./classify.html#bus-frequency-los) article for more details.

```{r cache=FALSE, include=FALSE}
mapviewOptions(
  basemaps = c("OpenStreetMap", "CartoDB.Positron")
)
```


```{r, cache=FALSE}
mapview::mapview(
  lanes |> filter(is_bus_lane),
  layer.name="Bus lane",
  color="#3BC1A8"
) + mapview::mapview(
  lanes |> filter(!is_bus_lane & frequency>=5 & !is.na(n_lanes) & n_lanes_direction>1),
  layer.name="NO bus lane with 5 or + bus/h + 1 lane/dir",
  color="#F63049"
)
```
