---
title: "Prioritize bus lane implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prioritize bus lane implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  cache = TRUE,
  comment = "#>"
)
```

```{r setup, cache=FALSE}
library(GTFShift)
library(tidytransit)
library(mapview)
library(dplyr)
library(osmdata)
```


# Introduction

Bus lanes have the potential to improve the reliability of bus operations by limiting the negative impacts of traffic congestion, reducing the variability of travel times, and increasing the average commercial speed - which can ultimately be used to increase service frequency.

However, introducing them on road infrastructure affects other modes, such as private vehicles, which may experience a decrease in the level of service due to reduced allocated space, potentially jeopardizing public acceptance.

Common criteria for implementing bus lanes include:

-   **Frequency of buses (and trams) per hour and direction, at a peak hour;**
-   **Number of lanes in the same direction;**
-   Existing traffic conditions;
-   Existing bus lanes in the area (from a network continuity perspective).

GTFShift provides methods to analyse the dimensions in bold, namely:

-  `GTFShift::get_way_frequency_hourly()`, to obtain the frequency of services per hour for each road segment with transit service and the characteristics of such each segments (such as number os lanes).

> `GTFShift:: get_route_frequency_hourly()` is an alternative

- `GTFShift::osm_bus_lanes()`, to identify existing bus lanes in the road network.

This document explores how to use these methods in a combined way to assist public transport planners in prioritizing bus lane implementations. For details on the several encapsulated features, refer to the numerated articles in the menu, that explore in detail each of the specific approaches followed. 

# Prioritize lanes

`GTFShift::prioritize_lanes()` is a simple method that bundles the logic of the ones mentioned above, returning a data.frame with the relevant characteristics for each road segment with transit service. With a single call, it returns all the aggregated information needed to prioritize bus lane implementations.

```{r}
# Get GTFS from library GTFS database for Portugal
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id = "lisboa"
gtfs = GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers=FALSE)

osm_q = opq(bbox=sf::st_bbox(tidytransit::shapes_as_sf(gtfs$shapes)))  |>
  add_osm_feature(key = "route", value = c("bus", "tram")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)

lanes = prioritize_lanes(gtfs, osm_q)
summary(lanes)
```


The aggregated data can then be manipulated according to the prioritization criteria defined by the user. For instance, the following code classifies (in red) the road segments considering a minimum frequency of 5 buses/hour and more than 1 lane per direction as high priority for bus lane implementation. 

Additionally, it displays the current bus lane network, highlighting in green the segments that meet the frequency and lane criteria, and in yellow those that do not meet the criteria but still have bus lanes.

> The classification of frequent service as 5 buses/hour is based on the HCM guidelines for level of service, where a frequency of 5 buses/hour corresponds to a level of service C or better. Refer to the [Classify transit data](./classify.html#bus-frequency-los) article for more details.

```{r cache=FALSE, include=FALSE}
mapviewOptions(
  basemaps = c("OpenStreetMap", "CartoDB.Positron")
)
```


```{r, cache=FALSE}
mapview::mapview(
  lanes |> filter((frequency<=4 | (is.na(n_lanes) | n_lanes_direction<=1)) & is_bus_lane),
  layer.name="Bus lane with - 5 bus/h OR - 1 lane/dir",
  color="#DAD887"
) + mapview::mapview(
  lanes |> filter(frequency>=5 & !is.na(n_lanes) & n_lanes_direction>1 & is_bus_lane),
  layer.name="Bus lane with 5 or + bus/h + 1 lane/dir",
  color="#3BC1A8"
) + mapview::mapview(
  lanes |> filter(frequency>=5 & !is.na(n_lanes) & n_lanes_direction>1 & !is_bus_lane),
  layer.name="NO bus lane with 5 or + bus/h + 1 lane/dir",
  color="#F63049"
)
```
