---
title: "7. GTFS Real Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{7. GTFS Real Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  cache = TRUE,
  comment = "#>"
)
```

```{r setup, cache=FALSE}
library(GTFShift)
library(dplyr)
library(mapview)
```

# Introduction

GTFS-RT extends the GTFS static data model by providing real time operational information. From service alerts, to trip updates, but also vehicle positions. The collection of this data can provide valuable insights about how planned operation performs in practice. `GTFShift` provides several methods to enable this data collection and analysis.


# Collect GTFS-RT data

To collect GTFS-RT data, use `GTFShift::rt_collect_json()` or `GTFShift::rt_collect_protobuf()`, depending on the encoding the feed uses (JSON or Protocol Buffers, respectively). These fetch GTFS-RT data from the endpoint and save it to a CSV file for further analysis. The method runs in a loop, fetching data at regular intervals (default is every 60 seconds) until manually stopped (CTRL+C).

```{r, eval=FALSE}
# Get GTFS from library GTFS database for Portugal
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id = "lisboa"

rt_collect_file <- "gtfs_rt_data.csv"
GTFShift::rt_collect_protobuf(data$URL.RT[data$ID == gtfs_id], rt_collect_file) # Run until manually stopped (CTRL+C)
```

# Extend prioritisation with speed from GTFS-RT data

Once GTFS-RT data is collected, it can be used to extend lane prioritisation analysis. 
`GTFShift::rt_extend_prioritisation()` takes a lane prioritisation data frame and a GTFS-RT collection (as an `sf` object) and enriches the prioritisation with real-time metrics. 
Refer to the method documentation for the full details.

```{r, eval=FALSE}
# Prioritisation based on static GTFS data and infrastructure characteristics
gtfs = GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers=FALSE)
osm_q = opq(bbox=sf::st_bbox(tidytransit::shapes_as_sf(gtfs$shapes)))  |>
  add_osm_feature(key = "route", value = c("bus", "tram")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)

lane_prioritisation <- GTFShift::prioritise_lanes(gtfs, osm_query)

# GTFS-RT data preparation
rt_collection = read.csv("rt_collect_file.csv") |> 
  sf::st_as_sf(coords = c("vehicle.position.longitude", "vehicle.position.latitude"), crs = 4326)

# (Optional) Filter updates, to remove those close to bus stops
gtfs_stops = tidytransit::stops_as_sf(gtfs$stops, crs=4326)
within_distance = st_is_within_distance(
  rt_collection |> st_transform(crs=3857), 
  gtfs_stops |> st_transform(crs=3857), 
  dist = stop_buffer_size
)
rt_collection_filtered = rt_collection[lengths(within_distance) == 0, ]

# Prioritisation extended with real-time data to add traffic conditions
lane_prioritisation_extended <- GTFShift::rt_extend_prioritisation(
  lane_prioritisation = lane_prioritisation,
  rt_collection = rt_collection_filtered
)
```

```{r include=FALSE}
lane_prioritisation_extended = sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.8.1/prioritization_lisboa_rt_gtfs2026-02-04_run20260226_extended.gpkg")
```

```{r cache=FALSE, include=FALSE}
mapviewOptions(
  basemaps = c("CartoDB.Positron")
)
```


The resulting `lane_prioritisation_extended` data frame includes additional columns with speed metrics, such as average speed, median speed, and speed percentiles, providing a more comprehensive view of lane performance based on real-time data.

```{r cache=FALSE}
lane_prioritisation_0800 = lane_prioritisation_extended |> filter(hour==8)

mapview::mapview(
  lane_prioritisation_0800,
  zcol = "speed_avg",
  layer.name = "Average speed per lane"
)

p50_frequency = quantile(lane_prioritisation_0800$frequency, 0.5, na.rm=TRUE)
p50_speed = quantile(lane_prioritisation_0800$speed_avg, 0.5, na.rm=TRUE)
mapview::mapview(
  lane_prioritisation_0800 |> filter(is_bus_lane & (frequency<p50_frequency | (is.na(n_lanes) | n_lanes_direction<=1) | speed_avg<=p50_speed)),
  layer.name=sprintf("Bus lane with -%d bus/h OR -2 lane/dir OR %.2f km/h or - avg. speed", p50_frequency, p50_speed),
  color="#DAD887",
  homebutton=FALSE,
  lwd=3

) + mapview::mapview(
  lane_prioritisation_0800 |> filter(is_bus_lane & frequency>=p50_frequency & !is.na(n_lanes) & n_lanes_direction>1 & speed_avg>p50_speed),
  layer.name=sprintf("Bus lane with +%d bus/h AND +1 lane/dir AND +%.2f km/h avg.speed", p50_frequency-1, p50_speed),
  color="#3BC1A8",
  homebutton=FALSE,
  lwd=3
) + mapview::mapview(
  lane_prioritisation_0800 |> filter(!is_bus_lane & frequency>=p50_frequency & !is.na(n_lanes) & n_lanes_direction>1 & speed_avg<=p50_speed),
  layer.name=sprintf("NO bus lane with +%d bus/h AND +1 lane/dir AND %.2f km/h or - avg.speed", p50_frequency-1, p50_speed),
  color="#F63049",
  homebutton=FALSE,
  lwd=3
)
```

## When speed information is missing: infer speed from position updates

If GTFS-RT does not provide speed information, it can be inferred from the progression of position updates through time.
The method `GTFShift::rt_average_speed()` implements this functionality, calculating, for each GTFS-RT update, the speed
based on the distance between the current and previous position (projected on the route geometry), 
divided by the time difference between the two updates. 

For more details, refer to the method documentation. The resulting speed can then be used to extend the prioritisation analysis as described above.
