---
title: "Prioritise bus lane implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prioritise bus lane implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  cache = TRUE,
  comment = "#>"
)
```

```{r setup, cache=FALSE}
library(GTFShift)
library(tidytransit)
library(mapview)
library(dplyr)
library(osmdata)
```

# Introduction

Bus lanes have the potential to improve the reliability of bus operations by limiting the negative impacts of traffic congestion, reducing the variability of travel times, and increasing the average commercial speed - which can ultimately be used to increase service frequency.

However, introducing them on road infrastructure affects other modes, such as private vehicles, which may experience a decrease in the level of service due to reduced allocated space, potentially jeopardizing public acceptance.

Common criteria for implementing bus lanes include:

-   Frequency of buses (and trams) per hour and direction, at a peak hour;
-   Number of lanes in the same direction;
-   Existing traffic conditions;
-   Existing bus lanes in the area (from a network continuity perspective).

GTFShift provides methods to analyse these dimensions, namely:

-  `GTFShift::get_way_frequency_hourly()`, to obtain the frequency of services per hour for each road segment with transit service and its associated characteristics (such as number of lanes).

- `GTFShift::osm_bus_lanes()`, to identify existing bus lanes in the road network.

- `GTFShift::rt_collect_json()` or `GTFShift::rt_collect_protobuf()`, to collect GTFS-RT data, which can be later used with `GTFShift::rt_extend_prioritisation()` to include real-time operational metrics in the prioritisation analysis.

This document explores how to use these methods in a combined way to assist public transport planners in prioritising bus lane implementations. For details on the several encapsulated features and method variations, refer to the numbered articles in the menu, that explore in detail each of the specific approaches followed. 

# Prioritise lanes

## Generate base indicators

`GTFShift::prioritise_lanes()` is a simple method that generates indicators for most of the criteria mentioned above using GTFS and OpenStreetMaps data (service frequency and lane characteristics). With a single call, it returns a data.frame with the relevant metrics for each road segment with transit service. 

```{r include=FALSE}
lanes <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.8.1/prioritization_lisboa_rt_gtfs2026-02-04_run20260226.gpkg")
```


```{r, eval=FALSE}
# Get GTFS from library GTFS database for Portugal
data <- read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))
gtfs_id <- "lisboa"
gtfs <- GTFShift::load_feed(data$URL[data$ID == gtfs_id], create_transfers = FALSE)

osm_q <- opq(bbox = sf::st_bbox(tidytransit::shapes_as_sf(gtfs$shapes))) |>
  add_osm_feature(key = "route", value = c("bus", "tram")) |>
  add_osm_feature(key = "network", value = "Carris", key_exact = TRUE)

lanes <- prioritise_lanes(gtfs, osm_q)
```

```{r}
summary(lanes)
```



## Extend with GTFS-RT data

If GTFS-RT data is available, it can be used to extend the prioritisation analysis with real-time operational metrics, such as average speed. This can help identify road segments where buses are experiencing significant delays due to traffic congestion, which may benefit from bus lane implementation.

```{r eval=FALSE}
rt_collection <- read.csv("rt_collect_file.csv") |>
  sf::st_as_sf(coords = c("vehicle.position.longitude", "vehicle.position.latitude"), crs = 4326)

lanes <- GTFShift::rt_extend_prioritisation(
  lane_prioritisation = lanes,
  rt_collection = rt_collection
)
```

```{r include=FALSE}
lanes <- sf::st_read("https://github.com/U-Shift/GTFShift/releases/download/v0.8.1/prioritization_lisboa_rt_gtfs2026-02-04_run20260226_extended.gpkg")
```

Refer to the [GTFS Real Time](./rt.html) article for details on how to collect GTFS-RT data and extend the prioritisation analysis.

## Analyze results

Method `GTFShift::get_prioritisation_stats()` can be used to obtain statistics about lane prioritisation, weighted by length and/or frequency.

```{r}
# For network analysis, frequency weight is more appropriate,
# to give more importance to the segments with more service
lanes_0800 <- lanes |> filter(hour == 8)
stats <- GTFShift::get_prioritisation_stats(lanes_0800, weight = "frequency")
stats

# At route level, length weight is more appropriate,
# to give more importance to the segments with higher extension,
# as the frequency does not vary along the route
lanes_0800_736 <- lanes |> filter(hour == 8 & grepl("199_0", routes))
stats_736 <- GTFShift::get_prioritisation_stats(lanes_0800_736, weight = "length")
stats_736
```

## Visualize results

The aggregated data can then be manipulated according to the prioritisation criteria defined by the user. For instance, the following code highlights (in red) the road segments as high priority for bus lane implementation if they have more than 1 lane per direction and a frequency above the median number of buses per hour registered at 8:00.


```{r}
lanes_0800 <- lanes |> filter(hour == 8)
summary(lanes_0800)

p50_frequency <- quantile(lanes_0800$frequency, 0.5, na.rm = TRUE)
p50_speed <- quantile(lanes_0800$speed_avg, 0.5, na.rm = TRUE)
```


```{r, cache=FALSE}
mapview::mapview(
  lanes_0800 |> filter(is_bus_lane & (frequency < p50_frequency | (is.na(n_lanes) | n_lanes_direction <= 1) | speed_avg <= p50_speed)),
  layer.name = sprintf("Bus lane with -%d bus/h OR -2 lane/dir OR %.2f km/h or - avg. speed", p50_frequency, p50_speed),
  color = "#DAD887",
  homebutton = FALSE,
  lwd = 3
) + mapview::mapview(
  lanes_0800 |> filter(is_bus_lane & frequency >= p50_frequency & !is.na(n_lanes) & n_lanes_direction > 1 & speed_avg > p50_speed),
  layer.name = sprintf("Bus lane with +%d bus/h AND +1 lane/dir AND +%.2f km/h avg.speed", p50_frequency - 1, p50_speed),
  color = "#3BC1A8",
  homebutton = FALSE,
  lwd = 3
) + mapview::mapview(
  lanes_0800 |> filter(!is_bus_lane & frequency >= p50_frequency & !is.na(n_lanes) & n_lanes_direction > 1 & speed_avg <= p50_speed),
  layer.name = sprintf("NO bus lane with +%d bus/h AND +1 lane/dir AND %.2f km/h or - avg.speed", p50_frequency - 1, p50_speed),
  color = "#F63049",
  homebutton = FALSE,
  lwd = 3
)
```

This visual representation allows to easily identify not only the high-priority segments for bus lane implementation, but also their spatial distribution across the existent network. A process that extends the results by incorporating the network continuity perspective, enabling the identification and eventual prioritisation of critical segments that connect bus lanes but have a bad performance.
