---
title: "3. Aggregate GTFS feeds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Aggregate GTFS feeds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.leaflet-container {
  width: 100% !important;
  min-width: 0 !important;
  box-sizing: border-box;
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, 
  message = FALSE,
  cache = TRUE,
  comment = "#>"
)
```

```{r setup, cache=FALSE}
library(GTFShift)
library(tidytransit)
library(dplyr)
library(mapview)
mapviewOptions(basemaps = c("Esri.WorldGrayCanvas", "CartoDB.Positron", "OpenStreetMap")) # Avoid Carto API Key
```

# Introduction 

Public transit analysis takes advantage of the standardized GTFS format. However, its provision by operator makes it difficult for network aggregated analysis, considering connectivity and multimodality. `GTFShift::unify()` proposes a simple solution to this problem, generating an aggregated GTFS file given several instances of these.

> This article uses GTFS feeds from the library GTFS database for Portugal as an example. Refer to the [vignette("download")](./download) for more details.



```{r eval=FALSE}
# Get GTFS from library GTFS database for Portugal
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))

gtfs_list = lapply(c("barreiro", "fertagus"), function(ID) {
  feed = GTFShift::load_feed(data$URL[data$ID == ID], create_transfers=FALSE)
  summary(feed)
  return(feed)
})
```

```{r echo=FALSE}
# FIx for https://github.com/U-Shift/GTFShift/issues/81
# Get GTFS from library GTFS database for Portugal
data = read.csv(system.file("extdata", "gtfs_sources_pt.csv", package = "GTFShift"))

gtfs_list = lapply(c("barreiro", "fertagus"), function(ID) {
  feed = GTFShift::load_feed(data$URL[data$ID == ID], create_transfers=FALSE)
   # If feed has columns start_pickup_drop_off_window and end_pickup_drop_off_window, remove them (they are not supported by gtfsrouter)
  if ("start_pickup_drop_off_window" %in% names(feed$stop_times)) {
    feed$stop_times <- feed$stop_times |> select(-start_pickup_drop_off_window, -end_pickup_drop_off_window)
  }
  summary(feed)
  return(feed)
})
```


# Unify GTFS

The unification is performed through `GTFShift::unify()`, producing a single GTFS instance, saved as a ZIP file. Option `create_transfers` enables the generation of `transfers.txt`, aggregating close stops, even if from different GTFS. 

```{r}
# Perform unification
gtfs_united = GTFShift::unify(gtfs_list[[1]], gtfs_list[[2]], create_transfers = TRUE)

summary(gtfs_united)

summary(gtfs_united$transfers)
```

It can be displayed using `mapview`.

<div style="display: flex; gap: 20px; overflow-x: auto;">
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Aggregated routes

```{r}
shape_agency = gtfs_united$trips |>
  left_join(gtfs_united$routes, by = "route_id") |>
  left_join(gtfs_united$agency, by = "agency_id") |>
  select(shape_id, agency_id, agency_name) |>
  distinct()

shapes_sf = tidytransit::shapes_as_sf(gtfs_united$shapes) |>
  left_join(shape_agency, by = "shape_id") |>
  filter(!is.na(agency_id))
```
```{r, cache=FALSE}
mapview::mapview(shapes_sf, zcol = "agency_name", legend = TRUE, layer.name="Agency")
```

</div>
<div style="flex: 1 1 0; min-width: 288px; max-width: 100%;">

### Aggregated stops

```{r}
stop_agency = gtfs_united$stop_times  |>
  left_join(gtfs_united$trips, by = "trip_id") |>
  left_join(gtfs_united$routes, by = "route_id") |>
  left_join(gtfs_united$agency, by = "agency_id") |>
  select(stop_id, agency_id, agency_name) |>
  distinct()

stops_sf = tidytransit::stops_as_sf(gtfs_united$stops) |>
  left_join(stop_agency, by = "stop_id") |>
  filter(!is.na(agency_id))
```
```{r, cache=FALSE}
mapview::mapview(stops_sf, zcol = "agency_name", legend = TRUE, layer.name="Agency")
```

</div>
</div>


